Mining of Keystroke and Mouse Dynamics to Increase the Engagement of Students with Programming Assignments

### 2018-04-21 paper.tex
Changes to conclusions
will leave issues open but I am addressing issue #12.

\begin{table*}[!t]
\centering
\caption{Performance of every classifier with 10-fold cross validation \cite{ijcci17}. Accuracy ($\kappa$) }
\label{tab_performance}
    \begin{tabular}{ | c | l | l | l | l | }
    \hline
    Affect          & Na\"ive Bayes           & J-48                & k-NN              & ANN \\
    \hline
    Flow/engaged    & 79.00\% (0.50) & 80.48\% (0.51) & 76.14\% (0.43) & 78.43\% (0.45) \\
    \hline
    Relaxation      & 71.10\% (0.40) & 72.57\% (0.45) & 69.14\% (0.37) & 71.24\% (0.34)\\
    \hline
    Distraction     & 70.33\% (0.43) & 71.00\% (0.43) & 74.00\% (0.50) & 72.52\% (0.39)\\
    \hline
    Frustration     & 74.71\% (0.49) & 73.86\% (0.45) & 72.62\% (0.42) & 78.00\% (0.50)\\
    \hline
    Boredom         & 74.71\% (0.47) & 84.57\% (0.59) & 83.81\% (0.58) & 76.19\% (0.45)\\
    \hline
    \end{tabular}
\end{table*}

\begin{table*}[!t]
\centering
\caption{Performance of classifiers for Flow/Engaged  with 10-fold cross validation. }
\label{tab_perf_flow}
    \begin{tabular}{ | l | l | l | }
    \hline
    Classifier   &  Accuracy            & $\kappa$                 \\
    \hline
    Na\"ive Bayes     & 72.52\%  & 0.339  \\
    \hline
    Deep Neural Network     & 78.29\%  & 0.419  \\
    \hline
    Random Forest      & 78.81\%  & 0.360  \\
    \hline
    Gradient Boosted Trees     & 78.95\%  & 0.415  \\
    \hline
    KNN      & 76.19\%  & 0.242  \\
    \hline
    Na\"ive Bayes Kernel     & 80.22\%  & 0.346 \\
    \hline
    J48     & 78.19 \%  & 0.433  \\
    \hline

@@

Table \ref{tab_perf_flow} shows the performance for the additional classifiers
evaluated specifically for the Flow/engaged affective state. The goal was to
increase the performance achieved earlier. Table 1shows the performance for the
additional classifiers evaluated specifically for the Flow/engaged affective
state. The goal was to increase the performance achieved earlier.  Again a ten-
fold cross-validation was conducted. In this case, instead of using an
evolutionary algorithm to conduct feature selection, a vector of weights was
optimized. The feature-weight vector assignes a weight to each feature. The
parameters used for the evolutionary optimization algorithm are the following:
Population size= 10, generations= 30, selection= tournament, tournament size=
0.25 (fraction of the population),  initialize probability=0.5,  mutation
probability = 1/number of attributes,  crossover probability=0.5. Parameters
used for the Random Forest algorithm: number of trees=10, criterion=gain_ratio,
maximal depth=20, confidence=0.25, minimal gain=0.1, minimal leaf size=2.  The
parameters used for are the J48 are the following:  confidence=0.79, minimal
leaf size=2. For the kNN algorithm, the value k=4 was used. For gradient boosted
trees: number of trees= 20,  maximal depth=5, minimum rows=10, minimum split
improvement=0.0, number of bins=20, learning rate=0.1, sample rate=1.0.  Bayes
Kernel:  laplace correction=True, estimation mode=greedy, minimum bandwidth=0.1,
number of kernels=10.  Finally for deep learning: activation=Rectifier, hidden
layer sizes= [50, 50], epochs=10. Again the accuracy and $\kappa$ values of the
algorithms show the difficulty of having classes that are not uniformly
distributed, with several algorithms reaching an acceptable accuracy but with a
low value of  $\kappa$, this is an indication of algorithms classifying almost
everything as the dominant class. The results of the tree induction algorithms
are better when we consider both accuracy and  $\kappa$ values. In the case of
kernel-based or deep learning methods they provide competitive results but at a
higher computational cost.

@@

The proposed method in this work obtained
satisfactory results. It is usual for a classification method based on free-text
to obtain an accuracy and $\kappa$ measure below their counterparts of
fixed-text; however, in this
work the results obtained were similar to those works applied to fixed-text
dynamics. A possible explanation of this would be the addition of the mouse
dynamics features and the additional preprocessing performed on the
feature vectors. Out of all methods tested, a multilayer perceptron
trained with backpropagation, the one tagged ANN in RapidMiner,  gave the best classification
results.

While these are promising results further experiments are needed. Other experiments
should focus on novice programmers, but that is left as future
work. %This
% But you have to comment on how this addresses the problem. Which
% should be in the title that is not there yet - JJ
% You also have to draw a _hard_ conclusion. Is programming using free
% text better than what? Typing a fixed text?
% Is 80% effective?
% future work will focus on extracting other features from the learners\' interaction
% such as behavioral data, as proposed by Blikstein
% \cite{blikstein2011using}; the parameters proposed by this author include the size
% of the program and changes in the code. Other authors also propose treating the use
% of certain keys as features e.g., the delete key because is manly used for correcting
% sentences. Also a multi-modal approach is needed. Since all these need
% a completely new approach, a new methodology will be developed to
% gather data and take measures.

>>>>

\begin{table*}[!t]
\centering
\caption{Performance of every classifier with 10-fold cross validation \cite{ijcci17}. Each cell shows accuracy and ($\kappa$)}
\label{tab_performance}
    \begin{tabular}{ | c | l | l | l | l | }
    \hline
    Affect          & Na\"ive Bayes           & J-48                & k-NN              & ANN \\
    \hline
    Flow/engaged    & 79.00\% (0.50) & 80.48\% (0.51) & 76.14\% (0.43) & 78.43\% (0.45) \\
    \hline
    Relaxation      & 71.10\% (0.40) & 72.57\% (0.45) & 69.14\% (0.37) & 71.24\% (0.34)\\
    \hline
    Distraction     & 70.33\% (0.43) & 71.00\% (0.43) & 74.00\% (0.50) & 72.52\% (0.39)\\
    \hline
    Frustration     & 74.71\% (0.49) & 73.86\% (0.45) & 72.62\% (0.42) & 78.00\% (0.50)\\
    \hline
    Boredom         & 74.71\% (0.47) & 84.57\% (0.59) & 83.81\% (0.58) & 76.19\% (0.45)\\
    \hline
    \end{tabular}
\end{table*}

\begin{table*}[!t]
\centering
\caption{Performance of classifiers for Flow/Engaged  with 10-fold cross validation and optimized feature-weights }
\label{tab_perf_flow}
    \begin{tabular}{ | l | l | l | }
    \hline
    Classifier   &  Accuracy            & $\kappa$                 \\
    \hline
    Na\"ive Bayes     & 72.52\%  & 0.339  \\
    \hline
    Deep Neural Network     & 78.29\%  & 0.419  \\
    \hline
    Random Forest      & 78.81\%  & 0.360  \\
    \hline
    Gradient Boosted Trees     & 78.95\%  & 0.415  \\
    \hline
    KNN      & 76.19\%  & 0.242  \\
    \hline
    Na\"ive Bayes Kernel     & 80.22\%  & 0.346 \\
    \hline
    J48     & 78.19 \%  & 0.433  \\
    \hline

@@

Table \ref{tab_perf_flow} shows the performance for the additional classifiers
evaluated specifically for the Flow/engaged affective state. The goal was to
increase the performance achieved earlier. Again a ten-
fold cross-validation was conducted. In this case, instead of using an
evolutionary algorithm to optimize feature selection, it was used for 
optimizing a weight vector. 
The feature-weight vector assigned a weight to each o the features. The
parameters used for the evolutionary optimization algorithm are the following:
Population size = 10, generations = 30, selection = tournament, tournament size = 0.25 (fraction of the population),  initialize probability = 0.5,  mutation
probability = 1/number of attributes,  crossover probability = 0.5. Parameters
used for the Random Forest algorithm: number of trees = 10, 
criterion = gain ratio, maximal depth = 20, 
confidence = 0.25, minimal gain = 0.1, minimal leaf size = 2.  The
parameters used for are the J48 are the following:  confidence = 0.79, minimal
leaf size=2. For the kNN algorithm, the value k = 4 was used. 
For gradient boosted
trees: number of trees = 20,  maximal depth = 5, minimum rows = 10, 
minimum split 
improvement = 0.0, number of bins = 20, 
learning rate = 0.1, sample rate = 1.0.  Bayes
Kernel: laplace correction = True, estimation mode = greedy, 
minimum bandwidth = 0.1, number of kernels = 10.  Finally for deep learning: activation = Rectifier, hidden layer sizes = [50, 50], epochs = 10.

Again the accuracy and $\kappa$ values of the
algorithms show the difficulty of having classes that are not uniformly
distributed, with several algorithms reaching an acceptable accuracy but with a
low value of  $\kappa$, this is an indication of algorithms classifying almost
everything as the dominant class. The results of the tree induction algorithms
are better when we consider both accuracy and  $\kappa$ values. In the case of
kernel-based or deep learning methods they provide competitive results but at a
higher computational cost.

@@

The proposed method in this work obtained
satisfactory results. It is usual for a classification method based on free-text
to obtain an accuracy and $\kappa$ measure below their counterparts of
fixed-text; however, in this
work the results obtained were similar to those works applied to fixed-text
dynamics. A possible explanation of this would be the addition of the mouse
dynamics features and the additional preprocessing performed on the
feature vectors.  Out of all methods tested, tree induction algorithms
obtained better results when we take in to consideration both accuracy and  $\kappa$ values.

While these are promising results further experiments are needed. Other experiments should focus on novice programmers, but that is left as future
work.

Further experiments with more data and more representative examples for all  the
clasess should give researchers the hability to do stratified samples to balence
the class distribution and achieve better results. Also more experiments should
give more insigth on the relationship between affective state  and interactive
behavior. Results show near 80\% accuracy, these results  are competitive
against other classifiers of a similar domain, but we do not know yet what is
the effectiveness when applied to affective-aware learing environments.

%This
% But you have to comment on how this addresses the problem. Which
% should be in the title that is not there yet - JJ
% You also have to draw a _hard_ conclusion. Is programming using free
% text better than what? Typing a fixed text?
% Is 80% effective?
%  - done  Mario 
Future work will focus on extracting other features from the learners\' interaction such as behavioral data, as proposed by Blikstein
\cite{blikstein2011using}; the parameters proposed by this author include the size of the program and changes in the code. Other authors also propose treating the use of certain keys as features e.g., the delete key because is manly used for correcting sentences.
% Also a multi-modal approach is needed. Since all these need
% a completely new approach, a new methodology will be developed to
% gather data and take measures.
